---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
library(lubridate)
library(dplyr)
library(ggplot2)

unzip("activity.zip")
activity <- read.csv("activity.csv",stringsAsFactors=FALSE)
activity <- activity %>% mutate(date2=ymd(activity$date))

```

## What is mean total number of steps taken per day?

We create summary statistics from the interval data, first ignoring NA values
```{r}
stepStats<-activity %>% 
           filter(!is.na(steps)) %>%
           group_by(date) %>% 
           summarise(countPerDay=sum(steps))
```
Using this summary we can get a feeling for the distribution of total number of steps per day using a histogram
```{r}
qplot(stepStats$countPerDay, binwidth=1000)
```

The distribution appears to be mostly symmetrical. This is backed up by the fact the mean number of steps each day and 
median number of steps each day are very similar.

```{r}
mean(stepStats$countPerDay)
median(stepStats$countPerDay)
```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
